---
title: "exercise-10"
author: "Janny Mauricio Velasco Alban"
format: html
editor: visual
---

# Exercise 10

## Libraries

```{r}

library(tidyverse)
library(skimr)
library(broom)
library(infer)
library(car)
library(emmeans)
library(sjPlot)
```

## Dataset

```{r}

f<- "https://raw.githubusercontent.com/difiore/ada-datasets/refs/heads/main/AVONETdataset1.csv"

d<- read_csv(f, col_names = TRUE)

view(d)
names(d)
```

```{r}

d<-d%>%
  select(Species1, Family1, Order1, Beak.Length_Culmen, Beak.Width, Beak.Depth, Tarsus.Length, Wing.Length, Tail.Length, Mass, Habitat, Migration, Trophic.Level, Trophic.Niche, Min.Latitude, Max.Latitude, Centroid.Latitude, Primary.Lifestyle, Range.Size)

cat("Dimensions:", dim(d), "\n")
```

## Let's go to the analysis

### Skim

```{r}

# First lets skim the data

skim(d)


```

### Challenge 1: One-factor ANOVA 

#### Step 1

-   Make boxplots of log(**Mass**) in relation to **Trophic.Level** and **Migration** behavior type. For each plot, drop from the visualization all species records where the categorical variable of interest is missing from the dataset. Also, you will want to convert the variable **Migration** (which is scored as a number: “1”, “2”, or “3”) from class numeric to either being classified as a factor or as a character (string) variable.

```{r}

#Step 1

d<- d%>%
  mutate(Migration = as.factor(Migration)) #Migration to factor

# Boxplot: log(Mass) by Trophic Level

d %>%
  filter(!is.na(Trophic.Level))%>%
  ggplot(aes(x = Trophic.Level, y = log(Mass), fill = Trophic.Level))+
  geom_boxplot(alpha = 0.7, outlier.size = 0.8)+
  labs(
    title = "log(Mass) by Trophic Level",
    x = "Trophic Level", 
    y = "log(Mass)"
  )+
  theme_minimal()+
  theme(legend.position = "none")


```

```{r}

#Boxplot: log(Mass) ~ Migration (drop NAs in Migration)

d %>%
  filter(!is.na(Migration))%>%
  ggplot(aes(x = Migration, y = log(Mass), fill = Migration))+
  geom_boxplot(alpha = 0.7, outlier.size = 0.8)+
  labs(
    title = "log(Mass) by Migration",
    x = "Migration (1 = Sedentary, 2 = Partially Migratory, 3 = Migratory)", 
    y = "log(Mass)"
  )+
  theme_minimal()+
  theme(legend.position = "none")
```

#### Step 2

```{r}
# Step 2
## Run linear models using the lm() function to look at the relationship between log(Mass) and Trophic.Level and between log(Mass) and Migration.


mA<- lm(log(Mass) ~ Trophic.Level, data = d %>%
          filter(!is.na(Trophic.Level))
        )
summary(mA)

### Examine the output of the resultant linear models. Is log(Mass) associated with either Trophic.Level or Migration category? That is, in the global test of significance, is the F statistic large enough to reject the null hypothesis of an F value of zero?


mB<- lm(log(Mass) ~ Migration, data = d %>%
          filter(!is.na(Migration))
        )
summary(mB)

##Given the regression coefficients returned for your Migration model, which Migration categor(ies) are different than the reference level? What level is the reference level? Relevel and assess differences among the remaining pair of Migration categories.

###Reference to 2 to assess the 1 vs 3 comparison

d2 <- d %>%
  filter(!is.na(Migration)) %>%
  mutate(Migration = relevel(Migration, ref = "2"))

mB2 <- lm(log(Mass) ~ Migration, data = d2)
summary(mB2)

###Reference to 3

d3<- d %>%
  filter(!is.na(Migration)) %>%
  mutate(Migration = relevel(Migration, ref = "3"))

mB3 <- lm(log(Mass) ~ Migration, data = d3)
summary(mB3)


```

#### Step 3

Conduct a post-hoc Tukey Honest Significant Differences test to also evaluate which **Migration** categories differ “significantly” from one another

```{r}

#Step 3: Post-hoc Tukey HSD - Migration

mB_aov <- aov(log(Mass) ~ Migration, data = d %>%
                filter(!is.na(Migration)))
tukey_mig<- TukeyHSD(mB_aov, which = "Migration")
tukey_mig
plot(tukey_mig)+theme_minimal()


```

#### Step 4

Use a permutation approach to inference to generate a null distribution of F statistic values for the model of log(**Mass**) in relation to **Trophic.Level** and calculate a p value for your original F statistic

```{r}

#Step 4: Permutation-Based F Test - log(Mass) ~ Trophic.Level

d_trophic <- d %>%
  filter(!is.na(Trophic.Level), !is.na(Mass)) %>%
  mutate(logMass = log(Mass))

obs_F_trophic <- d_trophic %>%
  specify(logMass ~ Trophic.Level)%>%
  calculate(stat = "F")

cat("\nObserved F (logMass ~ Trophic.Level):", obs_F_trophic$stat, "\n")

##Null distribution per permutation (1000 reps)

set.seed(2026)
null_dist_trophic <- d_trophic %>%
  specify(logMass ~ Trophic.Level) %>%
  hypothesise(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "F")

##Permutation p-value

p_perm_trophic <- null_dist_trophic %>%
  get_p_value(obs_stat = obs_F_trophic, direction = "greater")

cat("Permutation p-value:", p_perm_trophic$p_value, "\n")

###Visualization

visualize(null_dist_trophic)+
  shade_p_value(obs_stat = obs_F_trophic, direction = "greater")+
  labs(
    title = "Permutation Null Distribution of F (logMass ~ Trophic.Level",
    x = "F Statistic", 
    y = "Count"
  )+theme_minimal()
```

### Challenge 2: Data Wrangling plus One- and Two-Factor ANOVA

#### Step 1

Create the following two new variables and add them to AVONET dataset:

```{r}

#Residuals of log(Beak.Length_Culmen) ~ log(Mass) and residuals of log(Tarsus.Length) ~ log(Mass)

d <- d%>%
  mutate(
    logMass = log(Mass), 
    logBeak = log(Beak.Length_Culmen),
    logTarsus = log(Tarsus.Length)
  )

## regression on complete cases

beak_model <- lm(logBeak ~ logMass, data = d, na.action = na.exclude)
tarsus_model <- lm(logTarsus ~ logMass, data = d, na.action = na.exclude)

## Residuals as new columns 

d<- d %>%
  mutate(
    relBeak = residuals(beak_model),
    relTarsus = residuals(tarsus_model)
  )

cat("\nNew variables created: relBeak, relTarsus\n")
glimpse(select(d, Species1, relBeak, relTarsus))

```

#### Step 2

Make a boxplot or violin plot of your new relative tarsus length variable in relation to **Primary.Lifestyle** and of your new relative beak length variable in relation to **Trophic.Niche**

```{r}

#Boxplots - relTrasus ~ Primary.Lifestyle, relBeak ~ Trophic.Niche

## Relative Tarsus ~ Primary.Lifestyle

d %>% 
  filter(!is.na(Primary.Lifestyle), !is.na(relTarsus)
         
         )%>%
  ggplot(aes(x = Primary.Lifestyle, y = relTarsus, fill = Primary.Lifestyle))+
  geom_violin(alpha = 0.5)+
  geom_boxplot(width =0.15, alpha=0.8, outlier.size = 0.5)+
  labs(
    title="Relative Tarsus Length by Primary Lifesttyle",
    x = "Primary Lifestyle",
    y = "Relative Tarsus Length (residual)"
  )+
  theme_minimal()+
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))


```

```{r}

## Relative Beak ~ Trophic.Niche

d %>%
  filter(!is.na(Trophic.Niche), !is.na(relBeak)) %>%
  ggplot(aes(x = Trophic.Niche, y = relBeak, fill = Trophic.Niche))+
  geom_violin(alpha = 0.5)+
  geom_boxplot(width =0.15, alpha=0.8, outlier.size = 0.5)+
  labs(
    title="Relative Beak Length by Trophic Niche",
    x = "Trophic Niche",
    y = "Relative Beak Length (residual)"
  )+
  theme_minimal()+
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

#### Step 3

Run ANOVA analyses to look at the association between geographic range size and the variable **Migration**.

```{r}

#ANOVA ~ log(Range.Size) ~ Migration

## Let's examine the distribution

hist(d$Range.Size, breaks = 100, main = "Range.Size distribution")+theme_minimal()
hist(log(d$Range.Size), breaks = 100, main = "Log(Range.Size) distribution")+theme_minimal()
```

```{r}

d <- d %>%
  mutate(logRS = log(Range.Size))

d_mig <- d %>%
  filter(!is.na(Migration), !is.na(logRS)) %>%
  mutate(Migration = droplevels(as.factor(Migration)))
```

```{r}
##Linear model: logRS ~ Migration (reference = "1)

mC <- lm(logRS ~ Migration, data = d_mig)
summary(mC)

## R2 = proportion of variance in logRS by Migration

cat("\nR2 (logRS ~ Migration):", round(summary(mC)$r.squared, 4), "\n")
```

```{r}

##Let's check the Migration2 and Migration3 coefficients and p-vales

d_mig2 <- d_mig %>% mutate(Migration = relevel(Migration, ref = "2"))
mC2 <- lm(logRS ~ Migration, data = d_mig2)
summary(mC2)

d_mig3 <- d_mig %>% mutate(Migration = relevel(Migration, ref = "3"))
mC3 <- lm(logRS ~ Migration, data = d_mig3)
summary(mC3)

## Post-hoc Tukey HSD

mC_aov <- aov(logRS ~ Migration, data = d_mig)
tukey_RS <- TukeyHSD(mC_aov, which = "Migration")
tukey_RS
plot(tukey_RS)+theme_minimal()

```

#### Step 4

Winnow your original data to just consider birds from the Infraorder “Passeriformes” (song birds).

```{r}

#Filter to Passeriformes; One-Factor ANOVAs for relBeak

pass <- d %>%
  filter(Order1 == "Passeriformes",
         !is.na(relBeak),
         !is.na(Primary.Lifestyle),
         !is.na(Trophic.Level)
    
  )

cat("\nPasseriformes subset:", nrow(pass), "species\n")
```

```{r}

## Boxplot: relBeak ~ Primary.Lifestyle

ggplot(pass, aes(x = Primary.Lifestyle, y = relBeak, fill = Primary.Lifestyle))+
  geom_boxplot(alpha = 0.7, outlier.size = 0.7)+
  labs(
    title = "Relative Beak Length ~ Primary Lifestyle (Passeriformes)",
    x = "Primary Lifestyle",
    y = "Relative Beak Length"
  )+
  theme_minimal()+
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))

```

```{r}

## Boxplot: relBeak ~ Trophic.Level

ggplot(pass, aes(x = Trophic.Level, y = relBeak, fill = Trophic.Level))+
  geom_boxplot(alpha = 0.7, outlier.size = 0.7)+
  labs(
    title = "Relative Beak Length ~ Trophic.Level (Passeriformes)",
    x = "Trophic.Level",
    y = "Relative Beak Length"
  )+
  theme_minimal()+
  theme(legend.position = "none",
        )
```

```{r}

## Boxplot: relBeak ~ Primary.Lifestyle × Trophic.Level (combination)

ggplot(pass, aes(x = Primary.Lifestyle, y = relBeak, fill = Trophic.Level))+
  geom_boxplot(alpha = 0.7, outlier.size = 0.5, position=position_dodge(0.8))+
  labs(
    title = "Relative Beak Length by Lifestyle x Trophic Level (Passeriformes)",
    x = "Primary Lifestyle",
    y = "Relative Beak Length",
    fill = "Trophic Level"
  )+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r}
## relBeak ~ Primary.Lifestyle

mD1<- lm(relBeak ~ Primary.Lifestyle, data = pass)
summary(mD1)

##relBeak ~ Trophic.Level

mD2<- lm(relBeak ~ Trophic.Level, data = pass)
summary(mD2)
```

#### Step 5

Run a two-factor model to look at the association between relative beak length and both **Primary.Lifestyle** and **Trophic.Level** among the passeriforms.

```{r}

#Two-Factor ANOVA (Additive) — relBeak ~ PL + TL

mE <- lm(relBeak ~ Primary.Lifestyle + Trophic.Level, data = pass)
summary(mE)

```

#### Step 6

Additional two-way model with the same dataset and predictors, but adding the possibility of an interaction term.

```{r}

# Two-Factor ANOVA (With Interaction) through colon operator

mF <- lm(
  relBeak ~ Primary.Lifestyle + Trophic.Level + Primary.Lifestyle:Trophic.Level,
  data = pass
)

summary(mF)

anova(mF)

## Compare model via AIC

AIC(mD1, mD2, mE, mF)

## Conclusion: The nD1 abd nD2 are the most accurate according to the AIC values. It means than using just one predictor we will get a better coefficient than using two. 
```

#### Step 7

Use the `interaction.plot()` and either the `plot_model()` or `emmip()` function to visualize the interaction between **Primary.Lifestyle** and **Trophic.Level**, based on both \[1\] the raw dataset and \[2\] your model that includes an interaction term.

```{r}

# Interaction Plots

## Base R interaction.plot() — raw dataset group means

interaction.plot(
  x.factor = pass$Primary.Lifestyle,
  trace.factor = pass$Trophic.Level,
  response = pass$relBeak,
  fun= mean,
  type= "b",
  col= c("steelblue", "tomato", "forestgreen", "goldenrod"),
  pch= 16,
  lwd= 2,
  xlab= "Primary Lifestyle",
  ylab= "Mean Relative Beak Length",
  main= "Interaction Plot (Raw Means): Lifestyle × Trophic Level",
  legend= TRUE,
  trace.label= "Trophic Level"
)+
  theme_minimal()
```

```{r}

## estimated marginal means with CIs

emmip(mF, Trophic.Level ~ Primary.Lifestyle, CIs = TRUE)+
  labs(
    title = "Estimated marginal means: Lifestyle x Trophic Level",
    x = "Primary Lifestyle",
    y = "EMM of Relative Beak Length"
  )+theme_minimal()
```

#### Step 8

Use this approach to see whether variances in across groups in your various models (e.g., for **relative beak length \~ trophic level**) are roughly equal. Additionally, do a visual check of whether observations and model residuals within groups look to be normally distributed.

```{r}

# Checking the assumptions

## Variance homogeneity: maxSD / minSD

### relBeak ~ Trophic.Level

sd_trophic <- pass %>%
  group_by(Trophic.Level)%>%
  summarise(n = n(), sd = sd(relBeak, na.rm = TRUE), groups = "drop"
            )
print(sd_trophic)
cat("SD ratio (max/min) ~ Trophic.Level:",
    round(max(sd_trophic$sd) / min(sd_trophic$sd), 3), "\n")


```

```{r}

## relBeak ~ Primary.Lifestyle

sd_lifestyle <- pass%>%
  group_by(Primary.Lifestyle)%>%
  summarise(n = n(), sd = sd(relBeak, na.rm = TRUE), groups = "drop")
print(sd_lifestyle)
cat("SD ratio (max/min) ~ Primary.Lifestyle:",
    round(max(sd_lifestyle$sd) / min(sd_lifestyle$sd), 3), "\n")
```

```{r}

## Levene's test

leveneTest(relBeak ~ Trophic.Level, data = pass)
leveneTest(relBeak ~ Primary.Lifestyle, data = pass)
```

```{r}

## Histograms to visualize 

pass%>%
  filter(!is.na(Trophic.Level)) %>%
  ggplot(aes(x = relBeak, fill = Trophic.Level))+
  geom_histogram(bins = 30, color = "white", alpha = 0.8)+
  facet_wrap(~ Trophic.Level, scales = "free_y")+
  labs(
    title = "Distribution of relBeak by Trophic Level",
    x = "Relative Beak Length",
    y = "Count"
  )+
  theme_minimal()+
  theme(legend.position = "none")


```

```{r}

## QQPlot within Trophic.Level by group

pass %>%
  filter(!is.na(Trophic.Level)) %>%
  ggplot(aes(sample = relBeak))+
  stat_qq()+
  stat_qq_line(color = "red")+
  facet_wrap(~ Trophic.Level)+
  labs(
    title = "QQ Plot of Relative Beak Length by Trophic Level",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  )+
  theme_minimal()
  


```

```{r}

## Residual diagnostic

par(mfrow = c(2,2))
plot(mF, main = "Diagnostics: relBeak ~ PL * TL (Passeriformes)")+theme_minimal()
par(mfrow = c(1,1))
```

```{r}

## Q-Q plot of model residuals

tibble(res = residuals(mF)) %>%
  ggplot(aes(sample=res))+
  stat_qq()+
  stat_qq_line(color= "red")+
  labs(
    title = "Q-Q plot of Residuals - Interaction Model (mF)",
    x = "Theoretical Quantiles",
    y = "Residuals"
  )+
  theme_minimal()

cat("\n Exercise 10 complete - all Challenges and Steps addressed. \n")
```

# The end!

All the best,

Janny Mauricio Velasco Alban
